---
title: "Homework 3 - 193DS Spring"
author: "Michael Torbensen"
format: pdf
date: May 26, 2026
execute: 
  message: FALSE
  warning: FALSE
---

```{r}
# attaching packages 

library(tidyverse)
library(janitor)
library(here)
library(effectsize)
library(readxl)
library (ggeffects)
library(rstatix)

# reading in kelp data 
kelp <- read.csv(
  here("data", "temp_kelp.csv"))

# reading in personal data 
my_data <- read.csv(
  here("data","Personal_data.csv"))

```

# Problem 1. Giant Kelp Fronds 

## a. An appropraite test 

To start off, in order to measure the strength and direction of the relationship between kelp length and temperature I will use Pearson's Correlation Coefficient. In order to visualize and model the relationship, I will use a simple linear regression. The Pearson's Correlation Coefficient provides a numerical value to describe the relationship between the two variables and the simple linear regression models this relationship with a line. 

## b. Create a visualization 

```{r}

# base layer: ggplot
ggplot(data = kelp, 
       # x-axis: temperature
       # y-axis: kelp elongation rate (cm/day)
       mapping = aes(x = temp_c,
                     y = kelp_elong)) +
  # first layer: adding data points 
  geom_point(color = "#2a9d8f",
             alpha = 0.7,
             size = 3) +
  # second layer: adding trend line with color and standard error
  geom_smooth(method = "lm",
              color = "#e76f51", 
              se = TRUE) +
  # labeling x- and y-axis 
  labs(
    x = "Temperature (°C)",
    y = "Frond Elongation Rate (cm/day)", 
    title = "Average Giant Kelp Frond Growth in Different Temperatures") +
  # adding different theme from default
  theme_minimal()

```


## c. Checking assumptions and running tests

```{r}

# Fit the model first to extract residuals ----
kelp_model <- lm(
  kelp_elong ~ temp_c,
  data = kelp)

# Check normality of residuals: Q-Q plot ----
qqnorm(residuals(kelp_model))
qqline(residuals(kelp_model),
       col = "#e76f51")

# Check homoscedasticity: residuals vs fitted values ----
plot(fitted(kelp_model),
     residuals(kelp_model),
     xlab = "Fitted values",
     ylab = "Residuals",
     pch = 19,
     col = "#2a9d8f")
abline(h = 0,
       col = "#e76f51",
       lty = 2)

```

```{r}

# Pearson's correlation coefficient test ----
cor.test(kelp$temp_c, 
         kelp$kelp_elong, 
         method = "pearson")

# Simple linear regression ----
summary(kelp_model)

```

Linearity was assessed visually using a scatterplot of temperature vs. elongation rate with a fitted regression line. Normality of residuals was checked using a Q-Q plot, and homoscedasticity was assessed using a residuals vs. fitted values plot. The scatterplot suggests a reasonably linear relationship, the Q-Q plot shows residuals falling close to the reference line, and the residuals vs. fitted plot shows no obvious pattern. All three assumptions appear reasonably satisfied too. 

## d. Results communication

To evaluate the strength of the relationship between temperature and giant kelp frond elongation rate, I used a Pearson's correlation coefficient and a simple linear regression. These tests were appropriate because both variables are continuous and the scatterplot suggested a linear relationship between them. 

Temperature was a significant negative predictor of giant kelp frond elongation rate (r = -0.69, p < 0.001 , R² = 0.473). As seawater temperature increased, kelp fronds grew more slowly, suggesting that warmer water conditions are affect Macrocystis pyrifera and inhibit its growth. This pattern is consistent with what we know about kelp ecology, giant kelp thrives in cold, nutrient-rich water, and elevated temperatures likely symbolize conditions such as reduced upwelling that limit the nutrients available for growth.

